Scaling rates of true polar wander in convecting planets and moons

### 2016-02-22 tpw_rate.tex
Address some comments

For mantle convection problems the moment of inertia tensor is commonly separated into three parts \citep{ricard1993polar}:
\begin{equation}
I_{ij}(t) = I_0 \delta_{ij} + J_{ij}(t) + E_{ij}(t)
\label{eq:separation}
\end{equation}
where $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
If we plug this decomposition into Equation~\eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
\begin{equation}
\mathbf{\Omega} \times (\mathbf{J} \cdot \mitbf{\Omega}) = -\mitbf{\Omega} \times ( \mathbf{E} \cdot \mitbf{\Omega} )
\label{eq:disequilibrium}
\end{equation}
This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches
of the convective part of the moment of inertia ($\mathbf{E}$) and the rotational deformation part of the moment of inertia ($\mathbf{J}$).
Our goal is to characterize this balance from a perspective of scaling and fluid dynamics.

@@

\begin{equation}
\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
\label{eq:energy}
\end{equation}
where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature
(see Table~\ref{tab:parameters} for definitions of the other parameters).
In addition we use the simple equation of state
\begin{equation}
\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right)
\label{eq:eos}
\end{equation}
Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it
(a fourth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
Convenient choices for these numbers are listed in Table \ref{tab:nondim}, along with approximate Earth-like values for them.

@@

\begin{table}
\centering
\caption{Parameters for rotating mantle convection}
\label{tab:parameters}
\begin{tabular}{@{}lcc}
Symbol & Definition\\
\hline
$R_i$ & inner radius \\
$R$ & outer radius \\
$\Omega_0$ & reference rotation rate \\
$\eta$ & viscosity \\
$\kappa$ & thermal diffusivity \\
$\alpha$ &  thermal expansivity \\ 
$g_0$ & reference gravity \\
$I_0$ & reference moment of inertia \\
$T_0$ & reference temperature \\
$\rho_0$ & reference density \\ 
$\Delta T$ & temperature drop across mantle
\end{tabular}
\end{table}

>>>>

For mantle convection problems the moment of inertia tensor is commonly separated into three parts \citep{sabadini1981pleistocene, spada1992excitation}:
\begin{equation}
I_{ij}(t) = I_0 \delta_{ij} + J_{ij}(t) + E_{ij}(t)
\label{eq:separation}
\end{equation}
where $I_0$ is the spherically symmetric reference moment, $J_{ij}$ is the contribution due to rotational deformation, and $E_{ij}$ is the contribution due to internal density anomalies, as well as the surface deflections caused by them.
If we plug this decomposition into Equation~\eqref{eq:quasistatic_liouville} the spherically symmetric part $I_0 \delta_{ij}$ drops out, and we are left with
\begin{equation}
\mathbf{\Omega} \times (\mathbf{J} \cdot \mitbf{\Omega}) = -\mitbf{\Omega} \times ( \mathbf{E} \cdot \mitbf{\Omega} )
\label{eq:disequilibrium}
\end{equation}
This form of the quasistatic Liouville equation makes clear that the polar wander problem represents a balance between the mismatches
of the convective part of the moment of inertia ($\mathbf{E}$) and the rotational deformation part of the moment of inertia ($\mathbf{J}$).
Our goal is to characterize this balance from a perspective of scaling and fluid dynamics.

@@

\begin{equation}
\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
\label{eq:energy}
\end{equation}
where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature
(he remaining parameters are defined in Table~\ref{tab:parameters}).
The gravity vector $\mathbf{g}$ is given by $\mathbf{g} = -\nabla V$, where the 
gravitational potential $V$ comes from solving Poisson's equation
\begin{equation}
\nabla^2 V = 4 \pi G \rho
\end{equation}.
For the purposes of scaling, we assume that gravity is the approximately constant
vector $g_0 \mathbf{\hat{r}}$ (as is the case for Earth's mantle).
In addition we use the simple equation of state
\begin{equation}
\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right)
\label{eq:eos}
\end{equation}
Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it
(a fourth one, defined by the ratio of the length of day to a diffusion timescale, does not appear in the governing equations).
Convenient choices for these numbers are listed in Table \ref{tab:nondim}, along with approximate Earth-like values for them.

@@

\begin{table}
\centering
\caption{Parameters for rotating mantle convection}
\label{tab:parameters}
\begin{tabular}{@{}lcc}
Symbol & Definition\\
\hline
$R_i$ & inner radius \\
$R$ & outer radius \\
$G$ & gravitational constant \\
$V$ & gravitational potential
$M$ & mass of the planet \\
$\Omega_0$ & reference rotation rate \\
$\eta$ & viscosity \\
$\kappa$ & thermal diffusivity \\
$\alpha$ &  thermal expansivity \\ 
$g_0$ & reference gravity \\
$I_0$ & reference moment of inertia \\
$T_0$ & reference temperature \\
$\rho_0$ & reference density \\ 
$\Delta T$ & temperature drop across mantle
\end{tabular}
\end{table}
